---
title: "Build Prediction Model"
author: "Minghe Wang, Zebang Zhang(zz3309)"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(reshape2)
```

# Exploratory Data Analysis

```{r data loading}
load("./data/dat1.RData")
load("./data/dat2.RData")

# no missing data
all(is.na(dat1))
all(is.na(dat2))

ifelse(all(names(dat1) == names(dat2)), "train and test data have same structure", "train and test data have different structure")
str(dat1)
```
## Univariate analysis(continous & categorical)

```{r EDA_univar}
continuous_var <- dat1 %>% 
  select(age, height, weight, bmi, SBP, LDL, time, log_antibody)

categorical_var <- dat1 %>% 
  select(gender, race, smoking, diabetes, hypertension) %>% 
  mutate(
    # Convert binary variables to factors with labels
    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")),
    diabetes = factor(diabetes, levels = c(0, 1), labels = c("No", "Yes")),
    hypertension = factor(hypertension, levels = c(0, 1), labels = c("No", "Yes"))
  )

# Continuous:
summary(continuous_var)
# Boxplots
continuous_var_long <- continuous_var %>%
  tidyr::pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(continuous_var_long, aes(x = variable, y = value)) +
  geom_boxplot(fill = "lightblue") +
  facet_wrap(~variable, scales = "free", ncol = 3) +
  theme_minimal() +
  labs(title = "Boxplots of Continuous Variables")

# Categorical:
summary(continuous_var)
# bar plots
categorical_var_long <- categorical_var %>%
  tidyr::pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(categorical_var_long, aes(x = value)) +
  geom_bar(fill = "salmon") +
  facet_wrap(~variable, scales = "free", ncol = 3) +
  theme_minimal() +
  labs(title = "Bar Plots of Categorical Variables", x = "Category", y = "Count")
```
According to the box plot for continuous variables:

- `Age`, `BMI`, and `SBP` appear reasonably normally distributed, with expected ranges for an adult population; `LDL` cholesterol and `time` since vaccination show a wider range and some outliers, which may impact linear models.

- `log_antibody` (response) appears fairly symmetrical, which supports its use as a continuous response in linear or GAM models.

- Correlations and non-linear trends should be assessed in the next step to guide model form.

According to the bar plot for categorical variables:

- `Gender` is fairly balanced between Female and Male; 

- `Race` is skewed, with a majority of participants identifying as White (Category 1). Other racial/ethnic groups are underrepresented; 

- `Smoking` status shows that the majority are never smokers (Category 0), with fewer current and former smokers; 

- A large proportion of participants do not have `diabetes`; 

- A moderate split exists for `hypertension`, which may contribute meaningfully to clinical outcome variation

- Demographically, the population is balanced by gender but skewed by race and smoking status.

Overall, we believe the response variable `log_antibody` is well-behaved, and further correlation analysis(eg. bivariate) is needed.



```{r}
ggplot(dat1, aes(x = log_antibody)) +
  geom_density(fill = "skyblue", alpha = 0.5) +
  ggtitle("Density plot of log_antibody") +
  xlab("log_antibody") +
  theme_minimal()

ggplot(dat1, aes(x = log_antibody)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black", alpha = 0.7) +
  ggtitle("Histogram of log_antibody") +
  xlab("log_antibody") +
  theme_minimal()

```


```{r}
categorical_vars <- c("gender", "race", "smoking", "diabetes", "hypertension")
continuous_vars <- c("age", "height", "weight", "bmi", "SBP", "LDL", "time")
```


```{r}
# continous variable
continuous_vars <- c("age", "height", "weight", "bmi", "SBP", "LDL", "time")

for (var in continuous_vars) {
  corr <- cor(dat1[[var]], dat1$log_antibody, use = "complete.obs")
  p <- ggplot(dat1, aes_string(x = var, y = "log_antibody")) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    ggtitle(paste("Scatterplot of", var, "\nCorrelation:", round(corr, 3))) +
    theme_minimal()
  print(p)
}

```

```{r}
# categorical data
categorical_vars <- c("gender", "race", "smoking", "diabetes", "hypertension")
dat1[categorical_vars] <- lapply(dat1[categorical_vars], factor)

for (var in categorical_vars) {
  p <- ggplot(dat1, aes_string(x = var, y = "log_antibody", fill = var)) +
    geom_boxplot() +
    ggtitle(paste("Boxplot of log_antibody by", var)) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}

```

```{r}
dat_cont <- dat1[ , continuous_vars]

# coefficient matrix
cor_matrix <- cor(dat_cont, use = "complete.obs", method = "pearson")

print(round(cor_matrix, 2))

```

```{r}
cor_melt <- melt(cor_matrix)

ggplot(cor_melt, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1), space = "Lab",
                       name = "Pearson\nCorrelation") +
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1,
                                   size = 12, hjust = 1)) +
  coord_fixed() +
  ggtitle("Correlation Heatmap of Continuous Variables")
```

```{r}
categorical_vars <- c("gender", "race", "smoking", "diabetes", "hypertension")

for (i in 1:(length(categorical_vars)-1)) {
  for (j in (i+1):length(categorical_vars)) {
    var1 <- categorical_vars[i]
    var2 <- categorical_vars[j]
    cat("\nContingency Table:", var1, "vs", var2, "\n")
    tab <- table(dat1[[var1]], dat1[[var2]])
    print(tab)
    cat("Chi-squared Test:\n")
    print(chisq.test(tab))
  }
}

```


# Model Training

```{r}
library(caret) 
library(mgcv)
library(pdp)
library(earth)
library(tidyverse)
library(ggplot2)
```


```{r}
ctrl1 <- trainControl(method = "cv", number = 5)

train_y <- dat1$log_antibody
train_x <- dat1[, -which(names(dat1) == "log_antibody")]


set.seed(2)
gam.fit <- train(train_x, train_y,
                 method = "gam",
                 # tuneGrid = data.frame(method = "GCV.Cp", select = c(TRUE,FALSE)),
                 trControl = ctrl1)

gam.fit$bestTune

gam.fit$finalModel
```

```{r}
mars_grid <- expand.grid(degree = 1:3, 
                         nprune = 2:15)

set.seed(2)
mars.fit <- train(train_x, train_y,
                  method = "earth",
                  tuneGrid = mars_grid,
                  trControl = ctrl1)

ggplot(mars.fit)

mars.fit$bestTune

coef(mars.fit$finalModel) 
```

```{r}
bwplot(resamples(list(mars = mars.fit,
                      gam = gam.fit)), 
                 metric = "RMSE")
```

```{r}
mars.pred <- predict(mars.fit, newdata = dat2)
# test RMSE
mars_test_rmse = sqrt(mean((mars.pred - dat2[, "log_antibody"])^2))


gam.pred <- predict(gam.fit, newdata = dat2)
# test RMSE
gam_test_rmse = sqrt(mean((gam.pred - dat2[, "log_antibody"])^2))
```

